Improving Item-Item Similarity Estimation during Collaborative Filtering

### 2014-03-25 paper/paper.tex
Adding to bib
Edited probabilistic model section for linear model

To keep the present work focused, we confine ourselves to memory-based
collaborative filtering (CF) in which we work with a logical user-item matrix.
Each element of the matrix corresponds to a rating given by a particular user to a
particular item. Further, we choose to limit our analysis to item-item
similarity but note that these techniques should be equally effective at
improving user-user similarity estimation. In this representaion, an item is
a vector of the ratings it has received from all users. We now formalize the 
notion of {\em true} item-item similarity. Two alternative models for the 
generation of user-item ratings are useful in arriving at two different 
definitions of the {\em true} item-item similarity.

@@

In the second model, we imagine an unlimited supply of users each of whom
chooses to rate a subset of items. The lack of a rating is now meaningful since
it indicates a lack of interest of a particular user for a particular item. This
is captured by the model through a default rating of zero \footnote{We address
the issue of rating bias and the precise meaning of a zero valued ratings later
in this report.}. At any point in time, we may estimate the true similarity in
this model of {\em preferential} rating, by computing the cosine similarity of
the two item vectors. Only in the limit of inifinite time does this estimate
converge to the true similarity, which we denote $TrueSimPref$.  That is,
\begin{align}
TrueSimPref(A, B) = lim_{t\to\infty}\frac{\sum\limits_{u\in U_{AB}}
r_{u,A}r_{u,B}}{\sqrt{\sum\limits_{u\in U_A} r_{u,A}^2}
\sqrt{\sum\limits_{u\in U_B} r_{u,B}^2}},
\end{align}
where $U_A$ is the set of users who have rated $A$, $U_B$ is the set of users
who have rated $B$, and $U_{AB}$ is the set of users who have rated both $A$ and
$B$.

@@

\section*{Model}

We model the problem probabilistically. Let $Y$ be a random variable
representing the $TrueSim$ of a randomly chosen pair of items. Let $X_n$ be a
random variable representing the observed $Sim$ of a pair of items with $n$
common users.

We model $Y$ as a Normal distribution 
\begin{align}
Y \sim N(\mu, \sigma_{1}^2)
\end{align}
where $\mu$ and $\sigma_{1}^2$ are the average similarity score and the variance
of similarity scores of all pairs of items, respectively. Since $X_n$ represents
a noisy reading of the true similarity $Y$, we model $X_n | Y$ as a Gaussian
error around $Y$:
\begin{align}
(X_n | Y=y) \sim N(y, \sigma_{2, n}^2)
\end{align}

Note that $\sigma_{2, n}^2$ represents how noisy the estimation of $Sim$ is when
there are $n$ common users. This should decrease as $n$ increases.

@@

In probabilistic terms, for an observed value $X_n$, we want to find the value
of $Y$ that most likely produced $X_n$. Thus we desire the Maximum Likelihood
Estimate of $Y | X_n$
\begin{align}
\argmax_yP(Y=y|X_n=x) &= \argmax_y\left[\frac{P(X_n=x |
Y=y)P(Y=y)}{P(X_n=x)}\right]
\\&= \argmax_y\left[P(X_n=x|Y=y)P(Y=y)\right]
\\&= 
\argmax_y\left[\frac{1}{\sigma_{2,n}\sqrt{2\pi}}\exp{\left(\frac{-(x-y)^2}
{2\sigma_{2,n}^2}\right)}
\frac{1}{\sigma_{1}\sqrt{2\pi}}\exp{\left(\frac{-(y-\mu)^2}
{2\sigma_{1}^2}\right)}\right]
\\&= \argmin_y\left[\frac{(x-y)^2}{2\sigma_{2,n}^2} +
\frac{(y-\mu)^2}{2\sigma_{1}^2}\right]
\\&= \argmin_y\left[\left(\sigma_{1}^2+\sigma_{2,n}^2\right)y^2 - 
2\left(\sigma_{1}^2x+\sigma_{2,n}^2\mu\right)y\right]
\end{align}
We find the exact minimum by taking the derivative with respect to $y$ and
setting it to zero:
\begin{align}
\frac{d}{dy}\left[\left(\sigma_{1}^2+\sigma_{2,n}^2\right)y^2 - 
2\left(\sigma_{1}^2x+\sigma_{2,n}^2\mu\right)y\right]
&= 2\left(\sigma_{1}^2+\sigma_{2,n}^2\right)y - 
2\left(\sigma_{1}^2x+\sigma_{2,n}^2\mu\right) 
\\&= 0
\end{align}
Solving for y:
\begin{align}
y &= \frac{\sigma_{1}^2x+\sigma_{2,n}^2\mu}{\sigma_{1}^2+\sigma_{2,n}^2}
\end{align}
which can be rewritten as:
\begin{align}
\left(y - \mu\right) &= \frac{\sigma_{1}^2}{\sigma_{1}^2+\sigma_{2,n}^2}
\left(x-\mu\right)
\end{align}

The model suggests that there is a linear correlation between $X_n$, the
observed similarity with $n$ common users, and $Y$, the true similarity.
Moreover, the best predicted $y$ is a linear combination of the mean and the
observed similiarity. Since $\sigma_{2,n}^2\ge0$, the slope is always $\le1$, 
meaning that the observed distance from $\mu$ (the average similarity) 
overestimates the true distance. This is reasonable, since there is a prior 
that expects the true similarity to be around $\mu$.

@@

\section*{Conclusion}

>>>>

To keep the present work focused, we confine ourselves to memory-based
collaborative filtering (CF) in which we work with a logical user-item matrix.
Each element of the matrix corresponds to a rating given by a particular user to a
particular item. Further, we choose to limit our analysis to item-item
similarity but note that these techniques should be equally effective at
improving user-user similarity estimation. In this representaion, an item is
a vector of the ratings it has received from all users. We now formalize the 
notion of item-item similarity. Two alternative models for the generation of 
user-item ratings are useful in arriving at two different definitions of the 
{\em true} item-item similarity.

@@

In the second model, we imagine an unlimited supply of users, each of whom
chooses to rate a subset of items. The lack of a rating is now meaningful since
it indicates a lack of interest of a particular user for a particular item. This
is captured by the model through a default rating of zero \footnote{We address
the issue of rating bias and the precise meaning of a zero valued ratings later
in this report.}. At any point in time, we may estimate the true similarity in
this model of {\em preferential} rating, by computing the cosine similarity of
the two item vectors. Only in the limit of inifinite time does this estimate
converge to the true similarity, which we denote $TrueSimPref$.  That is,
\begin{align}
TrueSimPref(A, B) = lim_{t\to\infty}\frac{\sum\limits_{u\in U_{AB}}
r_{u,A}r_{u,B}}{\sqrt{\sum\limits_{u\in U_A} r_{u,A}^2}
\sqrt{\sum\limits_{u\in U_B} r_{u,B}^2}},
\end{align}
where $U_A$ is the set of users who have rated $A$, $U_B$ is the set of users
who have rated $B$, and $U_{AB}$ is the set of users who have rated both $A$ and
$B$.

@@

\section*{Linear Approach}
\subsection*{Probabilistic Model}

In our first approach, we model true and observed similarity probabilistically. For
this analysis, we can consider an abstact $TrueSim$ which ignores the details of 
$TrueSimRand$ or $TrueSimPref$. Similarly, we consider $Sim_{n}$ and $GoldSim$ to be
the generic observed similarity scores. Let $Y$ be a random variable representing 
the $TrueSim$ of a randomly chosen pair of items. Let $X_{n}$ be a random variable 
representing the $Sim_{n}$ of a pair of items when $n$ common users are observed.

We model $Y$ as a Normal distribution 
\begin{align}
Y \sim N(\mu, \sigma_{1}^2)
\end{align}
where $\mu$ and $\sigma_{1}^2$ are the average similarity score and the variance 
of similarity scores of all pairs of items, respectively. Since $X_n$ represents
a noisy reading of the true similarity $Y$, we model $X_n | Y=y$ as a Gaussian
error around $y$:
\begin{align}
(X_n | Y=y) \sim N(y, \sigma_{2, n}^2)
\end{align}

Note that $\sigma_{2, n}^2$ represents how noisy $Sim_n$ is as an estimation of
$TrueSim$ and therefore should decrease as $n$ increases. The probability
distribution of $X_n$ can be visualized in Figure 1.

@@

The problem of approximating the $TrueSim$ from $Sim_n$ can now be seen as
finding the value of $Y$ that most likely produced $X_n$. In other words we 
desire the Maximum Likelihood Estimate of $Y | X_n$:
\begin{align}
\hat{y} &= \argmax_yP(Y=y|X_n=x) 
\\&= \argmax_y\left[\frac{P(X_n=x | Y=y)P(Y=y)}{P(X_n=x)}\right]
\\&= \argmax_y\left[P(X_n=x|Y=y)P(Y=y)\right]
\\&= 
\argmax_y\left[\frac{1}{\sigma_{2,n}\sqrt{2\pi}}\exp{\left(\frac{-(x-y)^2}
{2\sigma_{2,n}^2}\right)}
\frac{1}{\sigma_{1}\sqrt{2\pi}}\exp{\left(\frac{-(y-\mu)^2}
{2\sigma_{1}^2}\right)}\right]
\\&= \argmin_y\left[\frac{(x-y)^2}{2\sigma_{2,n}^2} +
\frac{(y-\mu)^2}{2\sigma_{1}^2}\right]
\\&= \argmin_y\left[\left(\sigma_{1}^2+\sigma_{2,n}^2\right)y^2 - 
2\left(\sigma_{1}^2x+\sigma_{2,n}^2\mu\right)y\right]
\end{align}
We find the exact minimum by taking the derivative with respect to $y$ and
setting it to zero:
\begin{align}
\frac{d}{dy}\left[\left(\sigma_{1}^2+\sigma_{2,n}^2\right)y^2 - 
2\left(\sigma_{1}^2x+\sigma_{2,n}^2\mu\right)y\right]
&= 2\left(\sigma_{1}^2+\sigma_{2,n}^2\right)y - 
2\left(\sigma_{1}^2x+\sigma_{2,n}^2\mu\right) 
\\&= 0
\end{align}
This produces the following linear equation for \hat{y} in terms of x:
\begin{align}
\hat{y} &= \frac{\sigma_{1}^2x+\sigma_{2,n}^2\mu}{\sigma_{1}^2+\sigma_{2,n}^2}
\end{align}
which can be rewritten as:
\begin{align}
\left(\hat{y} - \mu\right) &= \frac{\sigma_{1}^2}{\sigma_{1}^2+\sigma_{2,n}^2}
\left(x-\mu\right)
\end{align}

The model suggests that there is a linear correlation between $X_n$ ($Sim_n$)
and $Y$ ($TrueSim$). Moreover, the best predicted $TrueSim$ is a linear 
combination of the mean and the observed similiarity. Since $\sigma_{2,n}^2$ is 
nonnegative, the slope is always less than or equal to one, meaning that the 
observed distance from $\mu$ (the average true similarity) overestimates the 
true distance. This is reasonable, since there is a prior that expects the true 
similarity to be around $\mu$.

@@

\section*{Conclusion}
\begin{thebibliography}{9}

\bibitem{Su2009}
    Xiaoyuan Su and Taghi M. Khoshgoftaar.
    ``A Survey of Collaborative Filtering Techniques.''
    \emph{Advances in Artificial Intelligence.} vol. 2009,
    Article ID 421425, 19 pages, 2009. doi:10.1155/2009/421425
\bibitem{Breese1998}
    Breese, John S., David Heckerman, and Carl Kadie, 
    ``Empirical analysis of predictive algorithms for collaborative filtering.'' 
    \emph{Proceedings of the Fourteenth conference on Uncertainty in artificial intelligence.}
    Morgan Kaufmann Publishers Inc., 1998.
\bibitem{Lemire1998}
    Daniel Lemire and Anna Maclachlan.
    ``Slope One Predictors for Online Rating-Based Collaborative Filtering.''
    \emph{SDM.} Vol. 5, 2005.

\end{thebibliography}
